Structural Breaks in Inflation Dynamics within the European Monetary Union

### 2011-05-10 pkg/paper/Paper/draft.Rnw
corrected typo in appendix
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@81 b679b2f7-6539-45d0-a605-5ddcb976cde9
new draft
git-svn-id: svn://svn.r-forge.r-project.org/svnroot/glogis@84 b679b2f7-6539-45d0-a605-5ddcb976cde9

\Keywords{inflation rate, structural break, EMU, Generalized Logistic Distribution, euro}

@@

Ever since the beginning of the European Union, the topic of a common currency was a controversial issue. Although the Economic and Monetary Union (EMU) is now a fact, the discussion about the economic effects of the Euro is far from being settled. The controversial topics range from the question of whether or not the Eurozone is indeed an optimal currency area (OCA) all the way to the very survival of the Euro in light of the budgetary problems of some of its member states. The effects of monetary unions on a number of macroeconomic indicators, with inflation being the most important one in this context, is in the center of an ongoing debate: this regards the question about short-run and steady-state inflation uncertainty dealt with in \citet{cap}, or the structural convergence of the inflation rates in EU countries, which is the topic of an investigation by \citet{palomba}. \\

@@

%(...too short: need: methodological ideas and main questions..., value added: schiefe, schwere ränder)\\
% http://home.iitk.ac.in/~kundu/paper152.pdf  (gl-beschreibung in unterlagen), p.4 more peaked in center, but heavier tails
However, a deeper inquiry into this problem is in need of a method able to capture changes in the inflation dynamics. We contribute to this debate by developing a new method for testing for structural breaks in the dynamics of the inflation rates, which focuses on these changes and thereby upon the effect of the EMU on a number of European countries both within and without the Eurozone. The framework developed in this paper makes it possible to test for any significant changes of a countries inflation dynamics and thus helps interpreting the effects of the EMU. We focus on changes in the first three moments of these inflation rates, which we assume to be distributed according to a generalized logistic distribution (GL), as defined in Section \ref{s3}. The GL distribution is a suitable one due to its flexibility to capture nonsymmetry and because of its fatter tails. For this purpose, we developed a new estimation approach building on previous research by \citet{z07}. \\

@@

This framework allows us to observe the evolution of mean, variance and skewness over time. 

@@

@@

\clearpage
\section{Data}\label{s2}

We employ seasonally unadjusted HICP series of 21 countries ranging from January 1990 to December 2010 that are obtained from the \citet{oecd}. Countries included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta are not included since they are very small economies. \\
The countries in this sample can be divided into three different groups. First the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and Spain).  Second, the EU members not participating in the ERM~II: Czech Republic, Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU. \\
%(... schlussverkäufe schon hier erwähnen, einfach cool, ja das gibts, wir haben die aber weggeschätzt; generell viel angelsächsischer/präpotenter schreiben; nur 1 weg der richtige und zwar unserer; nicht erwähnen, dass andere dinge auch noch möglich; so nach motto: mich interessierts, ich machs, andere erst am ende erwähnen...)
% greece: -for winter discounts: First 10 days of January-end of February; -for summer discounts: First 10 days of July -end od August; but has always been included
In the case of Belgium and Luxembourg, we dealt with the inclusion of the sales periods effects\footnote{This was demanded by a Commission regulation, details may be found in: \citet{eurlex} and in the Appendix \ref{dataprobs}.} into the HICP series (January 2000 for Belgium and January 1999 for Luxembourg) by estimating dummies for these month\footnote{There are two month with a significant decrease in prices (January and July), followed by significant price increases in the following months} and subtracting them from the original time series.

@@

To monitor changes in the dynamics of the inflation rates, we follow the usual strategy employed when testing for structural breaks. First, the model is fitted for the whole sample using the generalized logistic distribution. This then enables testing for the stability of the parameter estimates. If there is evidence of significant changes, the number and the timing of the break points are computed. \\
This framework allows to trace the evolution of mean, variance and skewness of the inflation series and thus interpreting deviations in any of these parameters as possibly due to a change in monetary policy, is made feasible. Any huge intervention or change is captured by fluctuations in the parameter stability of the series and can be quantified and related to a distinct time frame. \\
%(...discuss correlation...) \\
The correlation dynamics of the series is of less importance to this framework, which concentrates on changes in mean, variance and skewness and does not focus on autocorrelation, the use of alternative testes, as described in \cite{andreou}, is also hindered by the low number of observations, due to the monthly frequencies.\footnote{In some years the Billion Prices Project of the \cite{bpp}, which provides a much higher data frequency, could solve this problem.} 
%To provide consistent estimates of the parameter vector $\hat{\phi}$, the covariance matrix $\hat{V}^{-1/2}$ is estimated using a heteroskedasticity and autocorrelation consistent covariance estimator.

@@

with location $\theta$, scale $\sigma$ and shape $\delta$. For $\delta=1$ the distribution simplifies to the logistic distribution, for $\delta<1$ it is skewed to the left and for $\delta$$>$1 it is skewed to the right (see Appendix \ref{glscore} for more details). \\
In econometrics, the logistic distribution is often used in income distributions and growth models. This is due to  its fatter tails, which fit these kind of data somewhat better. \citet{won} employ a GL distribution in a regression model with autocorrelated errors. They use this type of distribution rather than a Student's t-distribution, as to model the fact that these are oftentimes  non symmetric and severely left or right skewed. A similar GL distribution is also used in \citet{tolikas} who analyze extreme risk and value--at--risk in the German stock market, although they do not use a shape parameter. Regarding inflation rates, the GL distribution is --- to our best knowledge --- only used in context with expected inflation. \\
\citet{batchelor} use a logistic distribution (not its generalization) to model the distribution of mean expected inflation rates. \\

@@

against the $H_1$ of changes in at least one of the parameters. We first estimate the parameters of the GL--distribution by means of maximum likelihood $\hat{\phi} ~=~ \underset{\phi}{argmax} \sum_{t=1}^n \log f(y_t|\phi)$. The parameter stability is assessed using the empirical scores $s(y_t|\hat{\phi})$ as measures of model deviation.\footnote{for details see Appendix \ref{glscore} and \citet{z07}}  The empirical scores process $\mathit{efp}(\cdot)$ captures deviations from a hypothesized zero mean over time and is given by:

\begin{eqnarray}
\mathit{efp}(t) ~=~ \hat{V}^{-1/2} n^{-1/2} \sum_{i=1}^{\lfloor nt \rfloor} s(y_i|\hat{\theta}, \hat{\sigma}, \hat{\delta}) \quad (0 \leq t \leq 1), \nonumber
\label{eqn2}
\end{eqnarray}

where $\hat{V}$ is some consistent estimator of the variance of the scores, which is used to decorrelate the series as to avoid problems due to serial correlation. Then a functional central limit theorem (FCLT) for $\mathit{efp}(\cdot)$ is employed, which converges to a 3-dimensional Brownian bridge:

@@

After the empirical fluctuation process is computed, it conveys information about departures from the $H_0$ of parameter stability over time. For a precise quantification of the actual deviation from its hypothesized zero--mean path, the main idea of the test statistics presented here is to aggregate the $\mathit{efp}(\cdot)$ in such a way as to be able to construct a test statistics so that critical values and p values can be derived. In this paper, we employ the Supremum of LM (supLM) test by \citet{andrews}. This test is well suited for single break alternatives and performs better if more than just one of the distribution parameters change.

@@

\subsection{Breakpoint Estimation}
%(...cite bai and perron 2003 paper here, nachlesen, findet achim super, bei slowenien will er noch konfidenzintervalle dazu...)\\
The dating of structural breaks builds upon the work of \cite{bp03}. If a break is detected via the supLM test, the next step is estimating the number and the timing of these break points. Breakpoints $\tau_1,...,\tau_B$ are estimated via maximization of the full segmented likelihood:

@@

If the test statistics is significant, at least one break is estimated. In principle, the optimal segmentation (i.e. the concrete dates of a break) can be computed once B (the number of breakpoints) is known. However, B needs to be based on the observed data as well. The solution is to compute the timing of the breaks for a number of different values for B and then choose B by means of an information criterion, in this case the LWZ (Liu, Wu and Zidek) as developed in \citet{liu}.
All parameters $\tau_1, \dots, \tau_B, \phi^{(1)}, \dots, \phi^{(B+1)}$ can be estimated jointly using dynamic programming.\footnote{More details for this estimation technique are provided in \cite{z10}.}

\clearpage
\section{An Example}\label{s4}
%(...more text: 1. motivation, 2.clear methodological steps, 3.interpretation...) \\
<<echo=FALSE>>=
        x <- results$Slovenia
        slovpval <- sctest(x$efp, functional = supLM(0.1))$p.value
@
To illustrate our method, we consider the case of Slovenia. After a period of relative instability following the independence from Yugoslavia, Slovenia was successfull in realigning its economy and introduced a number of reforms leading to a stable growth in the past years. The good economic performance made it possible for Slovenia to enter the ERM~II in June 2004 and later on to introduce the Euro in January 2007. The data collected contains the period from January 1996 onwards. In the first few years, Slovenia experienced very high inflation rates, but was successfull in reducing inflation to a much lower level. \\
A glance at the time series depicted in Figure \ref{fig1} of the monthly inflation rates reveals a decline in the mean around the middle of the observed time period. The $efp(\cdot)$ starts to significantly deviate as early as 2001 and reaches the peak somewhere between 2003 and 2004. The p--value of the supLM test is \Sexpr{round(slovpval, 5)} and clearly indicates at least one break.  The number of breaks is calculated by means of the LWZ criteria, which chooses one break in August 2003. In this year, Slovenia introduced a number of financial reforms and severely decreased the year on year growth in the M3 monetary aggregate. Combined, these actions were enough to ensure the successfull participation in the ERM~II a year later and the introduction of the Euro two years later.

@@

\newpage
\section{Results}\label{s5}

% das package colortbl (\usepackege[dvips]{colortbl}) ermöglicht z.B. mit
%\rowcolor[gray]{0.90} in einer Tabellenzeile diese farbig (bzw. hier grau) zu unterlegen.

\footnotesize
\begin{longtable}{|l|l|r|r|r|l|l|l|}
\hline
\multicolumn{8}{|c|}{Austria and Belgium} \\
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(1, 2), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The Eastern countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(3,5,10,15,17,18), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Italy and Spain} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(12, 19), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Ireland} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(11), format_moments)))
@
\hline
\multicolumn{8}{|c|}{No change countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(9,14,7,6), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The 90s} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(16,20, 21), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Former DM--zone} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(4,8,13), format_moments)))
@
\caption{Break point and parameter estimates and ERM information}
\label{resulttable}
\end{longtable}
% warning: the groups then change; Italy and Greece have not yet answered
% Italian Statistics: Your request has been correctly inserted in the system with the reference number 13078/2010.
% comment: the greek break is NOT sure, since they have not yet answered about sales periods

@@

%(...3 punkte darstellung noch einbauen...) \\
%(...andere struktur: 1.short introduction, 2. countries in detail (6.1), 3. summary (6.2) ...) \\
Table \ref{resulttable} presents the break dates and the parameter values of the segments as well as the dates of entry into the ERM/ERM~II and the date of the euro introduction. We formed seven groups of countries according to the observed pattern of break points. These seven groups consist of Austria and Belgium, the Eastern European countries, the two big southern European countries, Italy and Spain, Ireland, the countries with no change in the inflation dynamics, the countries with breaks in the early 90s and the member countries of the former DM--zone (excluding Belgium). \\

@@

The second group consists of the Eastern European countries: Czech Republic, Estonia, Hungary, Poland, Slovakia and Slovenia. In almost all of these countries, the mean and the variance of their inflation rates declined in the later part of the 90s (Slovenia is an exception with an increase in variance after the break). Most of them experienced a break in 1997/1998, with Poland, Slovakia and Slovenia somewhat later. The break during the late 90s is a result of these countries efforts to curb in inflation by decreasing money supply and indicates the time when these countries actually overcame the biggest transitionally shocks on their way towards free market economies. Slovakia is very similar to Slovenia, as its inflation dynamics too changed roughly one year prior to its entry into ERM~II. \\
Another combination of countries with rather identical results are the two biggest southern European countries, Italy and Spain. Both of them experienced high rates of inflation in the early 90s. Afterwards, in the phase leading up to the monetary union, mean and variance were greatly reduced. Soon after the fixing of the exchange rates, this trend changed significantly, leading to a higher mean and a much higher variance. A possible reason for this is that the fiscal responsibility that was required for an entry into the Eurozone was softened after this goal was achieved. This hypothesis gains some credibility considering the huge government deficits of these countries after 1999. \\

@@

Ireland can be considered a rather special case. It is the only economy where the effects of the ongoing financial crisis of 2008 are visible.  In Ireland, compared to other countries, the deflationary effect of the financial crises was more persistent. This deflationary trend, which for a time affected all countries, can be seen in Figure \ref{greyplot}, which depicts a rolling mean estimation, where the mean is calculated using the first moment of a fitted GL distribution. There is evidence of a decline in mean inflation, which returns to its previous path roughly a year later. The break in 2008 can be attributed to the strong contraction triggered by the bursting of the Irish housing bubble and a beginning deflation in concert with the massive costs of the bailout of the countries banking system. Surprisingly, this deflationary pressure is not visible in Spain, which was affected by a housing bubble of roughly the same size as Ireland.  \\

@@

The disparity of the national inflation rates is still an issue, many researchers, like \citet{hofmann}  find a considerable amount of inflation differentials. \citet{cap} estimate short-run and steady-state inflation uncertainty in 12 EMU countries and find a considerable degree of heterogeneity across EMU countries in terms of average inflation and its degree of persistence. In a paper examining structural convergence of the inflation rates in EU countries, \citet{palomba} try to answer the question if during the 1990s the inflation rate dynamics of EU countries become more similar. They find that convergence over time of inflation dynamics was only partly observable. 
In a paper studying core inflation and using an aggregated euro area inflation rate, \citet{morana} finds three regimes (roughly 1980---1984, 1984---1993 and 1993---2000) governing the core inflation rate.  \\
Concerning inflation differentials, Figure \ref{greyplot} depicts the inflation rates and a rolling mean estimation, where the mean is calculated using the first moment of a fitted GL distribution.\footnote{If more countries have the same inflation rate in a given month, the lines are overlapped and thus get darker. We wold see one black line if all countries would show the same month--on--month increase at any one time. The red lines depict the means.} We do see that many of these countries follow a rather similar path. One event clearly visible is the financial crises and the short deflationary phase starting somewhere around the autumn of 2007 and lasting a few months. Over the time, there is clear evidence, that the inflation rates converged towards some common -- albeit broad -- mean level.

@@

Regarding the general trend, Table \ref{resulttable} presents lower, medium and upper (25, 50, 75) percentiles of the first three moments of the inflation series during the convergence phase (January 1990 until December 1998), the Euro phase (January 1999 until August 2008) and the time after the financial crisis (since September 2008\footnote{After the insolvency of the Lehman Brothers in September 2008.} until now), for the countries in the ``core'' of the Eurozone (Austria, Belgium, Finland, France, Germany, Greece, Ireland, Italy, Luxembourg, Netherlands, Portugal and Spain) and all the others. \\
Regarding the mean, we see a clear evolution towards a lower mean in both groups, where the core countries show much lower values. Almost the opposite is true concerning the variance of the inflation rates of the core countries, that has increased to a much higher value than at the start of the sample, with the upper percentile during the 90s now almost equaling the lower percentile for the time after the financial crisis; the variance for the other countries has declined significantly. Concerning skewness, we see a clear trend in the core countries for skewness to become negative, indicating that the bulk of the values now lies right to the mean. The same is true -- although to a lesser extent -- for the other countries. \\

>>>>

\Keywords{inflation rate, structural break, EMU, Generalized Logistic Distribution 
}

@@

Ever since the beginning of the European Union, the topic of a common currency was a controversial issue. Although the Economic and Monetary Union (EMU) is now a fact, the discussion about the economic effects of the euro is far from being settled. The controversial topics range from the question of whether or not the eurozone is indeed an optimal currency area (OCA) all the way to the very survival of the euro in light of the budgetary problems of some of its member states. The effects of monetary unions on a number of macroeconomic indicators, with inflation being the most important one in this context, is in the center of an ongoing debate: this regards the question about short-run and steady-state inflation uncertainty (in context to inflation expectations) dealt with in \citet{cap}, or the structural convergence of the inflation rates in EU countries, which is the topic of an investigation by \citet{palomba}. \\
%(... will jesus früher anetua ...) \\

@@

However, a deeper inquiry into this problem is in need of a method able to capture changes in inflation dynamics. We contribute to this debate by developing a new method for testing for structural breaks in the dynamics of the inflation rates and thereby upon the effect of the EMU on a number of European countries both within and without the eurozone. The framework developed in this paper makes it possible to test for any significant changes of a countries inflation dynamics and thus helps interpreting the effects of the EMU. We focus on changes in the first three moments of these inflation rates, which we assume to be distributed according to a generalized logistic distribution (GL). The GL distribution is a suitable one due to its flexibility to capture nonsymmetry and because of its fatter tails. For this purpose, we develop a new estimation approach building on previous research by \citet{z07}. \\

@@

This framework allows us to assess the evolution of mean, variance and skewness over time. The skewness parameter turns out to be quite important to measure changes in the long run inflation dynamics, as it changes considerably over the time of the analysis. This trend would be all but invisible using a normal distribution.  

@@

%(... jesus: maybe comment on results here ...) \\

@@

\clearpage
\section{Data}\label{s2}
%(... first method, than data...) \\
We employ seasonally unadjusted HICP series of 21 countries ranging from January 1990 to December 2010 that are obtained from the \citet{oecd}. Countries included are Austria,	Belgium, Czech Republic,	Denmark, Estonia,	Finland,	France,	Germany,	Greece,	Hungary,	Ireland,	Italy,	Luxembourg,	Netherlands,	Poland,	Portugal,	Slovakia, Slovenia,  Spain,	Sweden, and	United Kingdom. Latvia and Lithuania as well as Bulgaria and Romania are excluded due to data scarcity. Cypria and Malta are not included since they are very small economies. \\
The countries in this sample can be divided into three different groups. First the euro countries (Austria, Belgium, Estonia, France, Finland, Germany, Greece, Ireland, Italy, Luxembourg, Netherlands, Portugal, Slovakia, Slovenia and Spain).  Second, the EU members not participating in the ERM~II: Czech Republic, Hungary, Poland, United Kingdom, and Sweden. Denmark stands on its own as a member of the EU and the ERM~II, but not yet a member of the EMU. \\
%(... schlussverkäufe schon hier erwähnen, einfach cool, ja das gibts, wir haben die aber weggeschätzt; generell viel angelsächsischer/präpotenter schreiben; nur 1 weg der richtige und zwar unserer; nicht erwähnen, dass andere dinge auch noch möglich; so nach motto: mich interessierts, ich machs, andere erst am ende erwähnen...)
% greece: -for winter discounts: First 10 days of January-end of February; -for summer discounts: First 10 days of July -end od August; but has always been included
As shown in a paper by \cite{pltm}, the harmonization of the treatment of sales prices led to considerable changes in some countries (Belgium, Italy, Luxembourg and Spain); failing to take this into account may lead to considerable distortions in the estimation. In these cases, we dealt with the inclusion of the sales periods effects\footnote{This was demanded by a Commission regulation, details may be found in: \citet{eurlex} and in the Appendix \ref{dataprobs}.} into the HICP series (January 2000 for Belgium, January 2001 for Italy, January 1999 for Luxembourg and January 2001 for Spain) by estimating dummies for these month\footnote{There are two month with a significant decrease in prices (January and July), followed by significant price increases in the following months (February and August).} and subtracting them from the original time series.

@@

To monitor changes in the dynamics of the inflation rates, we follow the usual strategy employed when testing for structural breaks. First, the model is fitted for the whole sample using the generalized logistic distribution. This then enables testing for the stability of the parameter estimates. If there is evidence of significant changes, the number and the timing of the break points are estimated. \\
This framework allows to trace the evolution of mean, variance and skewness of the inflation series and thus interpreting deviations in any of these parameters as possibly due to a change in monetary policy, is made feasible. Any huge intervention or change is captured by fluctuations in the parameter stability of the series and can be quantified and related to a distinct time frame. \\
%(...discuss correlation...) \\
The correlation dynamics of the series is of less importance to this framework, which concentrates on changes in mean, variance and skewness and does not focus on autocorrelation, the use of alternative tests, as described in \cite{andreou}, is also hindered by the low number of observations, due to the monthly frequencies.\footnote{In some years the Billion Prices Project of the \cite{bpp}, which provides a much higher data frequency, could solve this problem.} 
%To provide consistent estimates of the parameter vector $\hat{\phi}$, the covariance matrix $\hat{V}^{-1/2}$ is estimated using a heteroskedasticity and autocorrelation consistent covariance estimator.

@@

with location $\theta$, scale $\sigma$ and shape $\delta$. For $\delta=1$ the distribution simplifies to the logistic distribution, for $\delta<1$ it is skewed to the left and for $\delta$$>$1 it is skewed to the right (see Appendix \ref{glscore} for more details). \\
In econometrics, the logistic distribution is often used in income distributions and growth models. This is due to  its fatter tails, which fit these kind of data somewhat better. \citet{won} develop a maximum likelihood estimator if the underlying distribution is a GL distribution. A similar GL distribution is also used in \citet{tolikas} who analyze extreme risk and value--at--risk in the German stock market, although they do not use a shape parameter. Regarding inflation rates, the GL distribution is --- to our best knowledge --- only used in context with expected inflation. \\
\citet{batchelor} use a logistic distribution (not its generalization) to model the distribution of mean expected inflation rates. \\

@@

against the $H_1$ of changes in at least one of the parameters. We first estimate the parameters of the GL--distribution by means of maximum likelihood $\hat{\phi} ~=~ \underset{\phi}{argmax} \sum_{t=1}^n \log f(y_t|\phi)$. The parameter stability is assessed using the empirical scores $s(y_t|\hat{\phi})$ as measures of model deviation.\footnote{for details see Appendix \ref{glscore} and \citet{z07}}  The empirical fluctuation process $\mathit{efp}(\cdot)$ captures deviations from a hypothesized zero mean over time and is given by:

\begin{eqnarray}
\mathit{efp}(t) ~=~ \hat{V}^{-1/2} n^{-1/2} \sum_{i=1}^{\lfloor nt \rfloor} s(y_i|\hat{\theta}, \hat{\sigma}, \hat{\delta}) \quad (0 \leq t \leq 1), \nonumber
\label{eqn2}
\end{eqnarray}
%(... jesus: thinks this is incomplete ...) \\
where $\hat{V}$ is some consistent estimator of the variance of the scores, which is used to decorrelate the series as to avoid problems due to serial correlation. Then a functional central limit theorem (FCLT) for $\mathit{efp}(\cdot)$ is employed, which converges to a 3-dimensional Brownian bridge:

@@

After the $\mathit{efp}(\cdot)$ is computed, it conveys information about departures from the $H_0$ of parameter stability over time. For a precise quantification of the actual deviation from its hypothesized zero--mean path, the main idea of the test statistics presented here is to aggregate the $\mathit{efp}(\cdot)$ in such a way as to be able to construct a test statistics so that critical values and p values can be derived. In this paper, we employ the Supremum of LM (supLM) test by \citet{andrews}. This test is well suited for single break alternatives and performs better if more than just one of the distribution parameters change.

@@

\subsection{Breakpoint Estimation}

The dating of structural breaks builds upon the work of \cite{bp03}. If a break is detected via the supLM test, the next step is estimating the number and the timing of these break points. Breakpoints $\tau_1,...,\tau_B$ are estimated via maximization of the full segmented likelihood:

@@

If the test statistics is significant (in this case the supLM test), at least one break is estimated. In principle, the optimal segmentation (i.e. the concrete dates of a break) can be computed once B (the number of breakpoints) is known. However, B needs to be based on the observed data as well. The solution is to compute the timing of the breaks for a number of different values for B and then choose B by means of an information criterion, in this case the LWZ (Liu, Wu and Zidek) as developed in \citet{liu}.
All parameters $\tau_1, \dots, \tau_B, \phi^{(1)}, \dots, \phi^{(B+1)}$ can be estimated jointly using dynamic programming.\footnote{More details for this estimation technique are provided in \cite{z10}.}

\clearpage
\subsection{An Example}\label{s4}

<<echo=FALSE>>=
        x <- results$Slovenia
        slovpval <- sctest(x$efp, functional = supLM(0.1))$p.value
@
To illustrate our method, we consider the case of Slovenia. After a period of relative instability following the independence from Yugoslavia, Slovenia was successfull in realigning its economy and introduced a number of reforms leading to a stable growth in the past years. The good economic performance made it possible for Slovenia to enter the ERM~II in June 2004 and later on to introduce the euro in January 2007. The data collected contains the period from January 1996 onwards. In the first few years, Slovenia experienced very high inflation rates, but was successfull in reducing inflation to a much lower level. \\
A glance at the time series depicted in Figure \ref{fig1} of the monthly inflation rates reveals a decline in the mean around the middle of the observed time period. The $efp(\cdot)$ starts to cross the border as early as 2001 and reaches the peak somewhere between 2003 and 2004. The p--value of the supLM test is \Sexpr{round(slovpval, 5)} and clearly indicates at least one break.  The number of breaks is calculated by means of the LWZ criteria, which chooses one break in August 2003.\footnote{For the changes in the parameters, see Table \ref{resulttable}.} In this year, Slovenia introduced a number of financial reforms and severely decreased the year on year growth in the M3 monetary aggregate. Combined, these actions were enough to ensure the successfull participation in the ERM~II a year later and the introduction of the euro two years later.

@@

\newpage
\section{Results}\label{s5}
%(... jesus: show test statistic: no room...) \\
% das package colortbl (\usepackege[dvips]{colortbl}) ermöglicht z.B. mit
%\rowcolor[gray]{0.90} in einer Tabellenzeile diese farbig (bzw. hier grau) zu unterlegen.

\footnotesize
\begin{longtable}{|l|l|r|r|r|l|l|l|}
\hline
\multicolumn{8}{|c|}{Austria and Belgium} \\
\hline
Country & Segment & Mean & Variance & Skewness & ERM & ERM~II & Euro \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(1, 2), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The Eastern countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(3,5,10,15,17,18), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Italy and Spain} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(12, 19), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Ireland} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(11), format_moments)))
@
\hline
\multicolumn{8}{|c|}{No change countries} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(9,14,7,6), format_moments)))
@
\hline
\multicolumn{8}{|c|}{The 90s} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(16,20, 21), format_moments)))
@
\hline
\multicolumn{8}{|c|}{Former DM--zone} \\
\hline
<<echo=FALSE, results=tex>>=
writeLines(unlist(lapply(c(4,8,13), format_moments)))
@
\caption{Break point, parameter estimates and ERM information}
\label{resulttable}
\end{longtable}
% warning: the groups then change; Italy and Greece have not yet answered
% Italian Statistics: Your request has been correctly inserted in the system with the reference number 13078/2010.
% comment: the greek break is NOT sure, since they have not yet answered about sales periods

@@

%(...3 punkte darstellung noch einbauen...) \\
%(...andere struktur: 1.short introduction, 2. countries in detail (6.1), 3. summary (6.2) ...) \\
%(... jesus: opening confusing and obscure...)\\
Table \ref{resulttable} presents the break dates and the parameter values of the segments as well as the dates of entry into the ERM/ERM~II and the date of the euro introduction. We formed seven groups of countries according to the observed pattern of break points. These seven groups consist of Austria and Belgium, the Eastern European countries, the two big southern European countries, Italy and Spain, Ireland, the countries with no change in the inflation dynamics, the countries with breaks in the early 90s and the member countries of the former DM--zone (excluding Austria and Belgium). \\

@@

The second group consists of the Eastern European countries: Czech Republic, Estonia, Hungary, Poland, Slovakia and Slovenia. In almost all of these countries, the mean and the variance of their inflation rates declined in the later part of the 90s (Slovenia is an exception with an increase in variance after the break). Most of them experienced a break in 1997/1998, with Poland, Slovakia and Slovenia somewhat later. The break during the late 90s is a result of these countries efforts to curb in inflation by decreasing money supply and indicates the time when these countrie's actually overcame the biggest transitionally shocks on their way towards free market economies. Slovakia is very similar to Slovenia, as its inflation dynamics too changed roughly one year prior to its entry into ERM~II. \\
%(... jesus: maybe some reference on MP (what is MP???)...) \\
%(... thinks this is just data artefact, however even after desaisonalization the same effect...) \\
Another combination of countries with rather identical results are the two biggest southern European countries, Italy and Spain. Both of them experienced high rates of inflation in the early 90s. Afterwards, in the phase leading up to the monetary union, mean and variance were greatly reduced. Soon after the fixing of the exchange rates, this trend changed significantly, leading to a higher mean and a much higher variance. A possible reason for this is that the fiscal responsibility that was required for an entry into the eurozone was softened after this goal was achieved. \\

@@

Ireland can be considered a rather special case. 
%It is the only economy where the effects of the ongoing financial crisis of 2008 are visible.  
In Ireland, compared to other countries, the deflationary effect of the financial crises was more persistent. This deflationary trend, which for a time affected all countries, can be seen in Figure \ref{greyplot}, which depicts a rolling mean estimation, where the mean is calculated using the first moment of a fitted GL distribution. There is evidence of a decline in mean inflation, which returns to its previous path roughly a year later. The break in 2008 can be attributed to the strong contraction triggered by the bursting of the Irish housing bubble and a beginning deflation in concert with the massive costs of the bailout of the countries banking system. Surprisingly, this deflationary pressure is not visible in Spain, which was affected by a housing bubble of roughly the same size as Ireland, however in Spain housing prices did not adjust so abruptly and in such a degree.  \\

@@

<<echo=FALSE>>=
data("HICP", package = "glogis")
euhi <- HICP[, colnames(HICP) == "EU"]
euhi <- 100 * diff(log(na.omit(euhi)))
euhim1 <- mean(window(euhi, start=yearmon(1990), end= yearmon(1998+11/12)))
euhim2 <- mean(window(euhi, start=yearmon(1998), end=yearmon(2008+7/12)))
euhim3 <- mean(window(euhi, start=yearmon(2008+8/10), end=yearmon(2010+11/12)))
euhiv1 <- var(window(euhi, start=yearmon(1990), end= yearmon(1998+11/12)))
euhiv2 <- var(window(euhi, start=yearmon(1998), end=yearmon(2008+7/12)))
euhiv3 <- var(window(euhi, start=yearmon(2008+8/10), end=yearmon(2010+11/12)))
euhis1 <- skewness(window(euhi, start=yearmon(1990), end= yearmon(1998+11/12)))
euhis2 <- skewness(window(euhi, start=yearmon(1998), end=yearmon(2008+7/12)))
euhis3 <- skewness(window(euhi, start=yearmon(2008+8/10), end=yearmon(2010+11/12)))
@

The disparity of the national inflation rates is still an issue. Many researchers, like \citet{hofmann}  find a considerable amount of inflation differentials. \citet{cap} estimate short-run and steady-state inflation uncertainty in 12 EMU countries and find a considerable degree of heterogeneity across EMU countries in terms of average inflation and its degree of persistence. In a paper examining structural convergence of inflation rates in EU countries, \citet{palomba} try to answer the question if during the 1990s the inflation rate dynamics of EU countries become more similar. They find that convergence over time of inflation dynamics was only partly observable. 
In a paper studying core inflation and using an aggregated euro area inflation rate, \citet{morana} finds three regimes (roughly 1980---1984, 1984---1993 and 1993---2000) governing the core inflation rate.  \\
Concerning inflation differentials, Figure \ref{greyplot} depicts the inflation rates and a rolling mean estimation, where the mean is calculated using the first moment of a fitted GL distribution.\footnote{If more countries have the same inflation rate in a given month, the lines are overlapped and thus get darker. We wold see one black line if all countries would show the same month--on--month increase at any one time. The red lines depict the means.} We do see that many of these countries follow a rather similar path. One event clearly visible is the financial crises and the short deflationary phase starting somewhere around the autumn of 2007 and lasting a few months. Over time, there are hints, that the inflation rates converged towards some common -- albeit broad -- mean level. For the inflation series of the whole EMU, as published by the \citet{oecd}, we see an increase in mean from \Sexpr{round(euhim1, 5)} (1990(1)--1998(12)) to \Sexpr{round(euhim2, 5)} (1999(1)--2008(8)) with a strong decrease to \Sexpr{round(euhim3, 5)} (2008(9)--2010(12)) following the financial crises. Whereas variance has increased over time from    \Sexpr{round(euhiv1, 5)} to \Sexpr{round(euhim3, 5)}. Regarding skewness, we see a clear change towards a negative skewness from \Sexpr{round(euhis1, 5)} to \Sexpr{round(euhis3, 5)}.

%(... compare this with the emu inflation rate...)\\

@@

Regarding the general trend, Table \ref{grouptable} presents lower, medium and upper (25, 50, 75) percentiles of the first three moments of the inflation series during the convergence phase (January 1990 until December 1998), the euro phase (January 1999 until August 2008) and the time after the financial crisis (since September 2008\footnote{After the insolvency of the Lehman Brothers in September 2008.} until now), for the countries in the ``core'' of the eurozone (Austria, Belgium, Finland, France, Germany, Luxembourg and Netherlands), the PIIGS countries (Portugal, Italy, Ireland, Greece and Spain) and all the others. \\
%(... jesus would further differentiate between core and southeren core (or PIIGS)...)\\
